Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-31 manuscript/manuscript.Rmd
mini changes methods section, s_t -> N_t for states in the RL section

For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $S_t=(X_t,Y_t,Z_t)$.
At each time step, the agent observes the state $S$ and harvests $X$,
\begin{align*}
  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
\end{align*}
where the fish mortality-from-harvest $M_X=M_X(S_t)$ is the agent's action at time $t$. 
This secures a reward of $M_X(S_t)X$.
Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.

@@

- $T(s_{t+1}|s_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system,

- $r(s_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $s_t$,

- $d(s_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,

@@

At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | s_t)$.
In return, it receives a discounted reward $\gamma^t r(a_t, s_t)$.
The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.

With any fixed policy function, the agent will traverse a path 
$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
sampled randomly from the distribution
\begin{align*}
  p_\pi(\tau) = 
  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
\end{align*}
Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
$$  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
$$
where,
$$
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
$$
is the cumulative reward of path $\tau$.
The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.

@@

The policy function $\pi$ often lives in a high or even infinite-dimensional space.
This makes it unfeasible to directly optimize $\pi$.
In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions. <!-- ML: I am not following this line of thought here. I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.

@@

Since our state and observation spaces are continuous, we will focus on deep reinforcement learning throughout this paper.
Within the RL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
(See, e.g., [@franccois2018introduction] for an overview of different optimization schemes used in DRL.)

@@

\textbf{Training a DRL agent.}
We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $h_t = \pi_\theta(N_t)$, the harvest quota on species $X$ at that timestep.
Each episode in the training and in the evaluation stages was at most 200 time steps long.
Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
\begin{align}
  \label{eq:thresh}
  X_t \leq X_{\text{thresh.}},\quad
  Y_t \leq Y_{\text{thresh.}},\quad
  \text{or,}\quad
  Z \leq Z_{\text{thresh.}}.
\end{align}
The reward function had two componenets:
First, at each time step the harvest $h_t$ was received as a reward.
Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.

@@

\textbf{Parameter values used.}
The model's dynamic parameters were chosen as follows
\begin{align}
\label{eq:param vals}
\begin{split}
  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
\end{split}
\end{align}
Moreover, the variances for the stochastic terms were chosen as
$$
  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
$$
The threshold values used to determine an early episode end were chosen as
$$
  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
$$
where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies (for both the DRL and constant mortality approaches).

>>>>

For our use case, the environment will be a computational model of the population dynamics of a fishery, with the environment state at time $t$ being a vector of all the fish populations, $N_t=(X_t,Y_t,Z_t)$.
At each time step, the agent observes the state $S$ and harvests $X$,
\begin{align*}
  (X_t, Y_t, Z_t) \mapsto ((1-M_X)\cdot X_t, Y_t, Z_t),
\end{align*}
where the fish mortality-from-harvest $M_X=M_X(N_t)$ is the agent's action at time $t$. 
This secures a reward of $M_X(N_t)X$.
Afterwards, the environment undergoes a time step under its natural dynamics given by \eqref{eq:3d model stoch}.

@@

- $T(N_{t+1}|N_t, a_t)$: *transition operator*, a conditional distribution which describes the dynamics of the system (where $N_i\in\mathcal{S}$ are states of the environment),

- $r(N_t, a_t)$: *reward function*, the reward obtained after performing action $a_t\in\mathcal{A}$ in state $N_t$,

- $d(N_0)$: *initial state distribution*, the initial state of the environment is sampled from this distribution,

@@

At a time $t$, the MDP agent observes the full state $s_t$ of the environment and chooses an action based on this observation according to a *policy function* $\pi(a_t | N_t)$.
In return, it receives a discounted reward $\gamma^t r(a_t, N_t)$.
The discount factor helps regularize the agent, helping the optimization algorithm find solutions which pay off within a timescale of $t \sim \log(\gamma^{-1})^{-1}$.

With any fixed policy function, the agent will traverse a path 
$\tau=(N_0,\ a_0,\ N_1,\ a_1\ \dots,\ N_{t_{\text{fin.}}})$ 
sampled randomly from the distribution
\begin{align*}
  p_\pi(\tau) = 
  d(N_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | N_t ) T( N_{t+1} | N_t, a_t ).
\end{align*}
Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
$$  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
$$
where,
$$
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, N_t),
$$
is the cumulative reward of path $\tau$.
The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.

@@

The policy function $\pi$ often lives in a high or even infinite-dimensional space.
This makes it unfeasible to directly optimize $\pi$.
In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions.\footnote{
Policies are, in general, functions from state space to policy space.
In our case, these are real-valued functions on a cube $\pi:[0,2]^{\times 3}\to \mathbb{R}_+$.
The space of all such functions is highly singular, spanning a \emph{non-separable Hilbert space}.
Even restricting ourselves to continuous policy functions, we end up with a set of policies which span the infinite dimensional space $L^2([0,2]^{\times3})$.
One way to avoid optimizing over an infinite dimensional ambient space is to discretize state space into a set of bins.
This approach runs into tractability problems: First, the dimension of policy space scales exponentially with the number of species.
Second, even for a fixed number of species (e.g., 3), the dimension optimized over can be prohibitively large---for example if one uses 1000 bins for each population in a three-species model, the overall number of parameters being optimized over is $10^9$.
Neural networks with much smaller number of parameters, on the other hand, can be quite expressive and sufficient to find a rather good (if not optimal) policy function.
}
<!-- ML: I am not following this line of thought here. I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.

@@

We will focus on deep reinforcement learning throughout this paper.
Within the DRL literature there is a wealth of algorithms from which to choose from to optimize $\pi$, each with its pros and cons.
Most, if not all, of these are based on gradient ascent by using the technique of *back-propagation* to efficiently compute the gradient.
Here we have used only one such algorithm (*proximal policy optimization (PPO)*) for the sake of particularity---to draw a clear comparison between the RL-based and the classical fishery management approaches.
In practice, further improvements can be expected by a careful selection of the optimization algorithm. 
(See, e.g., [@franccois2018introduction] for an overview of different optimization schemes used in DRL.)

@@

\textbf{Training a DRL agent.}
We trained a DRL agent parametrized by a neural network $\theta$ with two hidden, 64-neuron, layers.
The state space used was a cube $[0,2]^{\times 3}$, while the policies explored mapped states $N_t$ to a single number, $M_X(N_t) = \pi_\theta(N_t)$, the harvest mortality on species $X$ at that timestep.
Each episode in the training and in the evaluation stages was at most 200 time steps long.
Episodes were ended early in case of a ‘‘near-extinction event,’’ a moment in which any of the populations dipped below a threshold value, i.e.,
\begin{align}
  \label{eq:thresh}
  X_t \leq X_{\text{thresh.}},\quad
  Y_t \leq Y_{\text{thresh.}},\quad
  \text{or,}\quad
  Z \leq Z_{\text{thresh.}}.
\end{align}
The reward function had two componenets:
First, at each time step the harvest $h_t$ was received as a reward.
Second, if the episode ended early (say, at time-step $t$) because of a near-extinction event, a negative reward of $-50/t$ was awarded.
The constant 50 was chosen so as to select against highly extractive policies in the early stages of training, and thus guide the agent towards sustainable policies.
This technique, falling under the ‘‘umbrella’’ category of *reward shaping*, is commonly used -- alongside hyper-parameter tuning and state space normalization -- to improve the speed of training by avoiding badly-performing local optima in the gradient ascent algorithm.

@@

\textbf{Parameter values used.}
The model's dynamic parameters were chosen as follows
\begin{align}
\label{eq:param vals}
\begin{split}
  r_X = K_X = r_Y = K_Y = 1, \quad \beta = 0.3, \quad c = 0.1\\
  c_{XY} = 0.5,\quad f = 0.15,\quad D = 1.1, \quad d_Z = 0.135.
\end{split}
\end{align}
Moreover, the variances for the stochastic terms were chosen as
$$
  \sigma^2(\eta_{X,t}) = \sigma^2(\eta_{Y,t}) = \sigma^2(\eta_{Z,t}) = 0.05.
$$
The threshold values used to determine an early episode end were chosen as
$$
  X_{\text{thresh.}} = 0.08,\quad Y_{\text{thresh.}} = Z_{\text{thresh.}} = 10^{-4},
$$
where the higher value of $X_{\text{thresh.}}$ helped select for more sustainable fishing strategies.
